---
title: "SEFIS Habitat Variable Reduction"
author: "KM Purcell"
date: "Monday, February 02, 2015"
output: word_document
---


## Habitat Classification

The *object of this analysis* is to simplify the 6-variable habitat description data from the SEFIS camera survey into a smaller set or univariate descriptor that is more amenable for downstream modeling applications.  This is essentially a dimensionality problem which are traditionally approached with **classification** or **ordination** based approaches.  

```{r data_input, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
library(dplyr)
library(vegan)
# Import data from Red Snapper for an example
#hab <- read.csv(file = file.choose(), header = TRUE, )
hab <- read.table(file = "C://Users//Kevin.Purcell//Documents//GitHub//lionfish//data//lionfish_video_data.csv", na.strings="N/A", sep=",", header=TRUE)
```

## Data Munging
Initial data was derived from an Access query, via Christina.  Nate sent an overview of the lionfish [field codes and metadata](https://mail.google.com/mail/u/0/#inbox/14b5003b047cb50e) for some context to the data vectors.


Initial issues with the imported data set:  

* The only designated NA strings are "N/A", the "unknown" and "other" strings are in fact information (re: convo w/ NB).  
* Missing values in frame vectors are being interpreted as NAs during import.  [ ]
* Time variable currently scored as a factor.                                   [ ]


First I want to start with the dimensionality reduction problem.  I will try to use a clustering approach.  So I will limit the data set to only those vectors relevant to habitat.


```{r data_munge}
# Limit to habitat variables
clust <- hab %>%
  filter(Year==2011) %>%
  select(Biotic_Density:Relief)
head(clust)
```
Now I want to remove any NAs from the habitat variables
```{r}
clust_input <- na.omit(clust)
summary(clust_input)
```

Have to convert factors to numerical values for interpretation by the `vegan()` functions.
```{r, echo=FALSE}
clust_input$codedBT[clust_input$Biotic_Type=="Algae"] <- "1"
clust_input$codedBT[clust_input$Biotic_Type=="Other"] <- "2"
clust_input$codedBT[clust_input$Biotic_Type=="Unknown"] <- "3"

clust_input$codedBH[clust_input$Biotic_Height=="High"] <- "1"
clust_input$codedBH[clust_input$Biotic_Height=="Low"] <- "2"

clust_input$codedS[clust_input$Size=="Coarse"] <- "1"
clust_input$codedS[clust_input$Size=="Continuous"] <- "2"

clust_input$codedR[clust_input$Relief=="High"] <- "1"
clust_input$codedR[clust_input$Relief=="Low"] <- "3"
clust_input$codedR[clust_input$Relief=="Moderate"] <- "2"
```
Re-coded variables look like:
```{r, echo=FALSE}
head(clust_input)
```



Now I want to convert that data.frame to a matrix for input with the `vegan()` functions
```{r}
clust_input2 <- clust_input %>%
  select(Biotic_Density, Substrate, codedBT:codedR)
summary(clust_input2)
clust_input2$codedBT <- as.numeric(clust_input2$codedBT)
clust_input2$codedBH <- as.numeric(clust_input2$codedBH)
clust_input2$codedS <- as.numeric(clust_input2$codedS)
clust_input2$codedR <- as.numeric(clust_input2$codedR)

# Convert to matrix for vegan
matrix <- as.matrix(clust_input2)
str(matrix)
```

## Analysis

Calculate a distance matrix using the `vegdist()` function.  An early reference to the use of [gower](https://stat.ethz.ch/pipermail/r-sig-ecology/2010-December/001791.html) dissimilarity indicated that the calculation was effective for use on both continuous and categorical variables.  

```{r dist_calc}
# Calculate the Gower Distance as an input of hclust
library(vegan)
distMatrix <- vegdist(matrix,method = "gower",na.rm = TRUE)
```
Use the Gower dissimilarity matrix as an input for the [hierarchical clustering](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/hclust.html) function in the `vegan` package.  

First, I will use the `"single"` clustering option, which is essentially a nearest-neighbor algorithm.  From our prospective this is comparable to a minimum spanning tree.  In this case I favor this approach because a strength of the `"single"` approach is its ability to locate discontinuities in the data which is our primary objective.  
```{r}
clus <- hclust(distMatrix, "single")
plot(clus)
```
Second I apply the `"comple"` option, which is a noted for is ability to form compact clusters, this is a secondary notion for us because it is has no real effect on our interpretation.  

```{r}
cluc <- hclust(distMatrix, "complete")
plot(cluc)
```

I attempted to use the `"average"` option, essentially because it was applied in my example approach and because it is considered a middle of the road approach.

```{r}
# This plot crashes R
clua <- hclust(distMatrix, "average")
plot(clua)
```

In my opinion for a theoretical stand point the initial `"single"` options fits our analysis criteria and shows solid ability to identify discontinuities in our actual input data. However to compare performance of the various linkage options I used the `cophenetic()` function which estimates the dissimilarity from each tree based on all pairs of points.  The name refers to **Cophenetic correlation** which measures the similarity between the original dissimilarity and estimated dissimilarity from the derived trees. 


```{r}
range(distMatrix)

cor(distMatrix, cophenetic(clus))

cor(distMatrix, cophenetic(cluc))

cor(distMatrix, cophenetic(clua))
```

The highest dissimilarity was derived from the `"average"` linkage approach so I will proceed with that.  
```{r}
plot(clua)
rect.hclust(clua, 3)
grp <- cutree(clua, 3)
```

A look at the distribution of substrate by groups
```{r, echo=FALSE}
par(mfrow=c(2,3))
boxplot(Biotic_Density ~ grp, data=matrix, notch=F, main="Biotic Density")
boxplot(Substrate ~ grp, data=matrix, notch=F, main="Substrate")
boxplot(codedBT ~ grp, data=matrix, main="Biotic Type")
boxplot(codedBH ~ grp, data=matrix, main="Biotic Height")
boxplot(codedS ~ grp, data=matrix, main="Size")
boxplot(codedR ~ grp, data=matrix, main="Relief")
```


Using an ordination approach we can look at the how our results compare between two distinct classification methodologies.  

```{r}
ord <- cca(matrix)
plot(ord, display="sites")
ordihull(ord, grp, lty = 2, col = "Blue")
```

Here there is considerable overlap among the groups, however with 6-dimensional data it is not necessarily a problem.  Primarily it corroborates what our Habitat-by-group boxplots showed, which was that each group was defined by a unique set of conditions not necessarily driven by on or two factors.  

```{r}
sessionInfo()
```

This will incorporate a *k-means* analysis

```{r}
library(reshape2)
library(ggplot2)

crit <- vector()
nk <- 1:10
for (k in nk){
  m <- kmeans(matrix, k)
  crit <- c(crit, sum(m$withinss))
}
crit

dsc <- data.frame(k=nk, crit=scale(crit))
dscm <- melt(dsc, id.vars="k", variable.name="Measure")

p <- ggplot(dscm, aes(x=k, y=value, colour=Measure))
p <- p + geom_point(aes(shape=Measure))
p <- p + geom_line(aes(linetype=Measure))
p <- p + scale_x_continuous(breaks=nk, labels=nk)
p <- p + theme(legend.position="none")
p
```

```{r}
nclust <- 4
model <- m.kms <- kmeans(scale(matrix), nclust)
dscm <- melt(model$centers)
names(dscm) <- c("Clusters", "Variable", "Value")
dscm$Cluster <- factor(dscm$Cluster)
dscm$Order <- as.vector(sapply(1:length(matrix), rep, nclust))
p <- ggplot(dscm,
            aes(x=reorder(Variable, Order),
                y=Value, group=Cluster, colour=Cluster))
p <- p + coord_polar()
p <- p + geom_point()
p <- p + geom_path()
p <- p + labs(x=NULL, y=NULL)
p <- p + theme(axis.ticks.y=element_blank(), axis.text.y=element_blank())
p
```

```{r}
source("http://onepager.togaware.com/CreateRadialPlot.R")
dsc <- data.frame(group=factor(1:4), model$centers)
CreateRadialPlot(dsc, grid.min=-2, grid.max=2, plot.extent.x=1.5)

#Viz a single group
CreateRadialPlot(subset(dsc, group==4), grid.min=-2, grid.max=2, plot.extent.x=1.5)

#multi-panel with ggplot
p1 <- CreateRadialPlot(subset(dsc, group==1), grid.min=-2, grid.max=2, plot.extent.x=1.5)
p2 <- CreateRadialPlot(subset(dsc, group==2), grid.min=-2, grid.max=2, plot.extent.x=1.5)
p3 <- CreateRadialPlot(subset(dsc, group==3), grid.min=-2, grid.max=2, plot.extent.x=1.5)
p4 <- CreateRadialPlot(subset(dsc, group==4), grid.min=-2, grid.max=2, plot.extent.x=1.5)

library(gridExtra)
grid.arrange(p1+ggtitle("Cluster1"), p2+ggtitle("Cluster2"),
             p3+ggtitle("Cluster3"), p4+ggtitle("Cluster4"))
```


